detrendr examples

### 2017-07-05 Notes/detrendify_spingarn.tex
Added reference on air quality.
Notes and timing comparison updates

\documentclass{article}
\usepackage[a4paper, top=3cm, bottom=3cm]{geometry}
\usepackage[latin1]{inputenc}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{tocloft}
\usepackage[usenames]{color}
\RequirePackage[OT1]{fontenc}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}

@@

We introduce a linear time algorithm for removing trends in time series.

Examples:
\begin{itemize}
\item Air quality 
\item ECG \cite{Sanyal2012}
\item Chromatogram baseline estimation  \cite{Ning2014, Ilewicz2015}
\item Galaxy spectrum baseline estimation \cite{Ilewicz2015,Bacher2016}
\item Identifying absorption dips from black body radiation
\end{itemize}

@@

@@

>>>>

\documentclass{article}
\usepackage[a4paper, top=3cm, bottom=3cm]{geometry}
\usepackage[latin1]{inputenc}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{tocloft}
\usepackage[usenames]{color}
\RequirePackage[OT1]{fontenc}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\graphicspath{{./Figures/}}

@@

\subsection{Background}

\textbf{Oh, Lee, and Nychka 2011 - Fast Nonparametric Quantile Regression With Arbitrary Smoothing Method} Unified framework for non-parametric quantile regression by approximating check loss function with quadratic loss near zero.\\
\textbf{Kim, Koh, Boyd, and Gorinevsky 2009} - Introduce the concept of l1 trend filtering.\\
\textbf{Tibshirani, 2014 - Adaptive piecewise polynomial estimation via trend filtering} Describes properties of trend filtering using quadratic loss. Shows that trend filtering estimates adapt to the local level of smoothness much better than smoothing splines, and exhibit
a remarkable similarity to locally adaptive regression splines. Prove that (with the right choice of tuning parameter) the trend filtering estimate
converges to the true underlying function at the minimax rate for functions whose kth derivative is of bounded variation.\\
\textbf{Ning, Selesnick, Duval 2014 - Chromatogram baseline estimation and denoising using sparsity (BEADS)} Address problem of estimating a smooth baseline in noisy data with drift. \\
We propose to use the trend filtering penalty with the check loss function to produce a non-parametric quantile regression estimate that can be computes using a linear time algorithm for removing trends in time series. Can be extended to model multiple quantiles and ensure non-crossing.

\subsection{Application}

	 \begin{figure}[!h] 
	 	\caption{Raw Data - three collocated sensors}
		\includegraphics[width = 5in]{spodRaw.png}
	\end{figure}
	 \begin{figure}[!h] 
	 	\caption{Detrended data - three collocated sensors}
	\includegraphics[width = 5in]{spodDetrended.png}
	\end{figure}

Examples:
\begin{itemize}
\item Air quality \cite{Apte2017}
\item ECG \cite{Sanyal2012}
\item Chromatogram baseline estimation  \cite{Ning2014, Ilewicz2015}
\item Galaxy spectrum baseline estimation \cite{Ilewicz2015,Bacher2016}
\item Identifying absorption dips from black body radiation
\end{itemize}

@@

\section{Related Work}

\begin{itemize}
	\item Quantile splines \cite{Oh2011}
	\item BEADS 
\end{itemize}

@@

\section{Numerical Studies}

We compare our detrending method with BEADS and the more general nonparametric quantile method introduced in  \cite{Oh2011}.

### 2017-07-05 Notes/timing_and_convergence_examples.Rmd
Notes and timing comparison updates

# Compare cholesky decompositions and solvers
```{r cholesky_comparison}
nSeq <- seq(500, 20000, 500) 
k <- 4

@@

@@

plot(Matrix_chol~n, timing, ylab="microseconds", 
     main="Matrix package sparse cholesky decomposition")
plot(chol_solve~n, timing, ylab ="microseconds", 
     main="RcppEigen sparse cholesky decomposition")
plot(chol_solve~n, timing, ylab = "microseconds", 
     main="My banded cholesky solver (black) vs. Matrix solver (blue)")
points(Matrix_solve~n, timing, col="blue")

@@

## Some examples of baseline fitting with detrendr

```{r simulate, include=FALSE, echo=FALSE}

simulate <- function(n, k){
  x <- seq(1/n, 1, length.out=n)
  f <- 2*(x + 2)^2 + 3*cos(3*pi*x)
  tau <- 5e4
  g1 <- 40*exp(-tau*(x-0.3)^2)
  g2 <- 30*exp(-tau/2*(x-0.58)^2)
  g4 <- 45*exp(-tau/7*(x-0.6)^2)
  g3 <- 37*exp(-tau*(x-0.63)^2)
  y <- f + g1 + g2 + g3 + g4 + rnorm(n)
  D <- get_Dk(n, k)
  M <- diag(n) + crossprod(D)
  cholM <- as.matrix(chol(M))
  return(list(y=y, f=f, x=x, D=D, M=M, cholM=cholM))
}
```

<!-- ## Current timing experiments -->
<!-- ```{r timing, include=FALSE, echo=FALSE, cache=TRUE, } -->
<!-- nSeq <- seq(100, 1500, 200) -->
<!-- timing <- data.frame(n=nSeq, R = NA, C=NA) -->
<!-- tau <- 0.01 -->
<!-- lambda <- 10 -->
<!-- step <- 1 -->

<!-- for (i in 1:length(nSeq)){ -->
<!--   simData <- simulate(nSeq[i], 3) -->
<!--   y <- simData$y -->
<!--   theta <- simData$y -->
<!--   D <- simData$D -->
<!--   eta <- matrix(D%*%theta) -->
<!--   M <- simData$M -->
<!--   cholM <- simData$cholM -->
<!--   timing$R[i] <- system.time(  -->
<!--       multi_step_R <- spingarn_multi_step_R(theta, eta, y, D, M, lambda,  -->
<!--                                             tau, step, 100))[3] -->
<!--   timing$C[i] <- system.time( -->
<!--     multi_step <- spingarn_multi_step(theta, eta, y, D, cholM, -->
<!--                                      lambda, tau, step, 100))[3] -->
<!-- } -->

<!-- plot(C~n, timing, type="l", col="blue", ylab="Elapsed Time (s) per 100 iterations", main = "R code is in black, C in blue") -->
<!-- lines(R~n, timing, type="l") -->
<!-- ``` -->

<!-- ## Convergence when n=100, lambda = 10 -->
<!-- ```{r convergence, include=TRUE, echo=TRUE, cache=TRUE} -->

<!-- n <- 500 -->
<!-- simData <- simulate(n, 3) -->

<!-- y <- simData$y -->
<!-- theta <- simData$y -->
<!-- D <- simData$D -->
<!-- eta <- matrix(D%*%theta) -->
<!-- M <- simData$M -->
<!-- cholM <- simData$cholM -->
<!-- tau <- 0.05 -->
<!-- lambda <- 10 -->
<!-- step <- 1 -->

<!-- max_iter <- 70000 -->
<!-- rerr <- double(max_iter-1) -->
<!-- THx <- matrix(NA, n, max_iter) -->
<!-- for (iter in 1:max_iter) { -->
<!--   one_step <- spingarn_one_step_R(theta, eta, y, D, M, lambda, tau, step) -->
<!--   theta <- one_step[[1]] -->
<!--   eta <- one_step[[2]] -->
<!--   THx[,iter] <- prox_f1(theta, y, tau) -->
<!--   if (iter > 1){ -->
<!--     rerr[iter] <- norm(as.matrix(THx[,iter] - THx[,iter-1]),'f')/ -->
<!--                     (1 + norm(THx[,iter,drop=FALSE], 'f')) -->
<!--   } -->
<!-- } -->
<!-- theta_last <- prox_f1(theta, y, tau) -->

<!-- par(mfrow=c(2,1), mar=c(3,4,1,1)) -->
<!-- plot(log(rerr), type="l") -->

<!-- plot(simData$x,simData$f,type='l',col='blue', -->
<!-- ylim=c(min(y),max(y)), lwd=3,ylab="", xlab="") -->
<!-- points(simData$x,y, cex=.6) -->
<!-- lines(simData$x,theta_last,col='red', lwd=3) -->

<!-- ``` -->

<!-- ## Convergence when n=1000, lambda = 100 -->
<!-- ```{r convergence2, include=TRUE, echo=FALSE, cache=TRUE} -->

<!-- n <- 1000 -->
<!-- simData <- simulate(n, 3) -->

<!-- y <- simData$y -->
<!-- theta <- simData$y -->
<!-- D <- simData$D -->
<!-- eta <- matrix(D%*%theta) -->
<!-- M <- simData$M -->
<!-- cholM <- simData$cholM -->
<!-- tau <- 0.05 -->
<!-- lambda <- 100 -->
<!-- step <- 1 -->

<!-- max_iter <- 70000 -->
<!-- rerr <- double(max_iter-1) -->
<!-- THx <- matrix(NA, n, max_iter) -->
<!-- for (iter in 1:max_iter) { -->
<!--   one_step <- spingarn_one_step_R(theta, eta, y, D, M, lambda, tau, step) -->
<!--   theta <- one_step[[1]] -->
<!--   eta <- one_step[[2]] -->
<!--   THx[,iter] <- prox_f1(theta, y, tau) -->
<!--   if (iter > 1){ -->
<!--     rerr[iter] <- norm(as.matrix(THx[,iter] - THx[,iter-1]),'f')/ -->
<!--                     (1 + norm(THx[,iter,drop=FALSE], 'f')) -->
<!--   } -->
<!-- } -->
<!-- theta_last <- prox_f1(theta, y, tau) -->

<!-- par(mfrow=c(2,1), mar=c(3,4,1,1)) -->
<!-- plot(log(rerr), type="l") -->

<!-- plot(simData$x,simData$f,type='l',col='blue', -->
<!-- ylim=c(min(y),max(y)), lwd=3,ylab="", xlab="") -->
<!-- points(simData$x,y, cex=.6) -->
<!-- lines(simData$x,theta_last,col='red', lwd=3) -->

<!-- ``` -->

<!-- ## Convergence when n=2000, lambda = 100 -->
<!-- ```{r convergence3, include=TRUE, echo=FALSE, cache=TRUE} -->

<!-- n <- 2000 -->
<!-- simData <- simulate(n, 3) -->

<!-- y <- simData$y -->
<!-- theta <- simData$y -->
<!-- D <- simData$D -->
<!-- eta <- matrix(D%*%theta) -->
<!-- M <- simData$M -->
<!-- cholM <- simData$cholM -->
<!-- tau <- 0.05 -->
<!-- lambda <- 100 -->
<!-- step <- 1 -->

<!-- max_iter <- 70000 -->
<!-- rerr <- double(max_iter-1) -->
<!-- THx <- matrix(NA, n, max_iter) -->
<!-- for (iter in 1:max_iter) { -->
<!--   one_step <- spingarn_one_step_R(theta, eta, y, D, M, lambda, tau, step) -->
<!--   theta <- one_step[[1]] -->
<!--   eta <- one_step[[2]] -->
<!--   THx[,iter] <- prox_f1(theta, y, tau) -->
<!--   if (iter > 1){ -->
<!--     rerr[iter] <- norm(as.matrix(THx[,iter] - THx[,iter-1]),'f')/ -->
<!--                     (1 + norm(THx[,iter,drop=FALSE], 'f')) -->
<!--   } -->
<!-- } -->
<!-- theta_last <- prox_f1(theta, y, tau) -->

<!-- par(mfrow=c(2,1), mar=c(3,4,1,1)) -->
<!-- plot(log(rerr), type="l") -->

<!-- plot(simData$x,simData$f,type='l',col='blue', -->
<!-- ylim=c(min(y),max(y)), lwd=3,ylab="", xlab="") -->
<!-- points(simData$x,y, cex=.6) -->
<!-- lines(simData$x,theta_last,col='red', lwd=3) -->

<!-- ``` -->

>>>>

# Compare cholesky decompositions and solvers as n increases
```{r cholesky_comparison, cache=TRUE}
nSeq <- seq(500, 50000, 500) 
k <- 3

@@

  cholM <- chol(M)
  timing$Matrix_solve[i] <- median(microbenchmark(
    x1 <- as.numeric(Matrix::solve(cholM, Matrix::solve(t(cholM), theta))))$time*1e-3)
  timing$chol_solve[i] <- median(microbenchmark(
    x2 <- chol_solve(cholM, chol_solve(t(cholM), theta, k, FALSE), 
                   k, TRUE))$time*1e-3)
  i <- i+1
}

t1 <- M%*%x1
plot(as.numeric(t1)~theta)

plot(chol_solve~n, timing, ylab = "microseconds", 
     main="My banded cholesky solver (black) vs. Matrix solver (blue)")
points(Matrix_solve~n, timing, col="blue")
abline(0, .12)

# Compare cholesky decompositions and solvers as k increases
```{r cholesky_comparison_k, cache=TRUE}
n <- 40
kSeq <- seq(2, 10, 1)

timing <- data.frame(k=kSeq, Matrix_chol = NA, Matrix_solve = NA, 
                     Eigen_chol = NA, chol_solve = NA)
i <- 1
for (k in kSeq){
  D <- get_Dk(n, k)
  M <-  as(Diagonal(n) + Matrix::crossprod(D), "dgCMatrix")
  theta <- rnorm(n)

@@

t1 <- M%*%x1
plot(as.numeric(t1)~theta)

plot(Matrix_chol~k, timing, ylab="microseconds", 
     main="Matrix package sparse cholesky decomposition")
plot(Eigen_chol~k, timing, ylab ="microseconds", 
     main="RcppEigen sparse cholesky decomposition")
plot(chol_solve~k, timing, ylab = "microseconds", 
     main="My banded cholesky solver (black) vs. Matrix solver (blue)")
points(Matrix_solve~k, timing, col="blue")

```
# Algorithm order
```{r algorithm order, cache=TRUE}
nSeq <- seq(500, 15000, 500) 
k <- 3

timing <- data.frame(n=nSeq, multistep = NA)
i <- 1
for (n in nSeq){
  D <- get_Dk(n, k)
  M <-  as(Diagonal(n) + Matrix::crossprod(D), "dgCMatrix")
  theta <- rnorm(n)
  lambda0 <- 2*10^(-6)
  lambda <- lambda0*n^(k-1)/factorial(k-1)
  y <- rnorm(n)
  eta <- matrix(D%*%theta)
  tau <- 0.05
  cholM <- chol(M)
  timing$multistep[i] <- median(microbenchmark(
      multi_step <- spingarn_multi_step(theta, eta, y, D, cholM,
                                 lambda, tau, 1, 100, k)
  )$time*1e-9)

  i <- i+1
  print(i)
}

plot(multistep~n, timing, ylab="Seconds", xlab = "n", main = "100 steps of algorithm")
abline(0, 2.95*1e-6)

@@
